---
title: "Using Internal cfa-dataops Data with nowcastNHSN"
output:
  rmarkdown::html_vignette:
    code_folding: show
vignette: >
  %\VignetteIndexEntry{Using Internal cfa-dataops Data with nowcastNHSN}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE, dev = "pdf", fig.keep = "none"}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "pdf",
  fig.keep = "none"
)
```

```{r setup}
library(nowcastNHSN)
library(baselinenowcast)
library(dplyr)
```

## Overview

CDC-internal users with a configured cfa-dataops Python environment and Azure access can use `cfa_dataops_source()` to fetch versioned NHSN HRD snapshots.
This is a special data path: it shells out to the `dataops_save` command at fetch time, so ordinary package users do not need cfa-dataops, Python catalog packages, or Azure credentials.

The source reads cfa-dataops version directories, maps each version date to the Saturday ending that MMWR epiweek, and uses that mapped date as `report_date`.
The returned data has the same reporting-triangle schema as the Delphi and hubData sources:

- `reference_date`
- `report_date`
- `location`
- `count`
- `signal`

## Creating a cfa-dataops Source

Install cfa-dataops and the public catalog in the workflow environment that will run this data source.
For example, using `uv`:

```{bash install-cfa-dataops, eval = FALSE}
uv venv /tmp/nowcastnhsn-cfa-dataops-venv --python 3.13
UV_CACHE_DIR=/tmp/uv-cache uv pip install \
  --python /tmp/nowcastnhsn-cfa-dataops-venv/bin/python \
  "cfa-dataops @ git+https://github.com/CDCgov/cfa-dataops.git@main" \
  "cfa-catalog-public @ git+https://github.com/CDCgov/cfa-catalog-pub.git"

export DATAOPS_SAVE=/tmp/nowcastnhsn-cfa-dataops-venv/bin/dataops_save
```

By default, `cfa_dataops_source()` uses the preliminary cfa-stf NHSN HRD dataset namespace, `public.stf.nhsn_hrd_prelim`.
Set `prelim = FALSE` to use `public.stf.nhsn_hrd`.

```{r create-source, eval = FALSE}
source <- cfa_dataops_source(
  target = "covid",
  command = Sys.getenv("DATAOPS_SAVE", "dataops_save")
)
```

The disease target maps to the NHSN HRD column names used by the cfa-stf pipeline:

- `target = "covid"` uses `totalconfc19newadm`
- `target = "flu"` uses `totalconfflunewadm`
- `target = "rsv"` uses `totalconfrsvnewadm`

## Fetching Reporting Data

The interface is the same `fetch_reporting_data()` generic used by the other data sources.
Pass Saturday week-ending `Date` vectors for `reference_dates` and `report_dates`.

```{r fetch-data, eval = FALSE}
reporting_data <- fetch_reporting_data(
  source = source,
  reference_dates = seq(
    as.Date("2025-08-09"),
    as.Date("2025-08-30"),
    by = "week"
  ),
  report_dates = seq(
    as.Date("2025-08-16"),
    as.Date("2025-09-06"),
    by = "week"
  ),
  locations = c("ca", "ny")
)
```

```{r fetch-data-fixture, include = FALSE}
# Load pre-computed fixture to avoid requiring cfa-dataops and Azure access on
# every vignette build. To regenerate: run the chunk above in a cfa-dataops
# enabled environment and saveRDS() the result to
# vignettes/cfa-dataops-source/reporting_data.rds #nolint
reporting_data <- readRDS("cfa-dataops-source/reporting_data.rds")
```

```{r show-data}
head(reporting_data)
```

Internally, the method calls `dataops_save` with a bounded version query and `--full_range`.
For the report dates above, the version query covers all cfa-dataops versions from the Sunday before the first report week through the day before the week after the last report week.

If multiple cfa-dataops versions fall in the same MMWR epiweek, `dedup = "latest"` keeps the most recent version for that `report_date`.
Use `dedup = "earliest"` to keep the first version in that report week.

```{r dedup, eval = FALSE}
reporting_data <- fetch_reporting_data(
  source = source,
  reference_dates = "*",
  report_dates = seq(
    as.Date("2025-08-16"),
    as.Date("2025-09-06"),
    by = "week"
  ),
  locations = "ca",
  dedup = "earliest"
)
```

## Reporting Triangle Check

The returned data can go through the same cumulative-to-incremental and reporting-triangle path as other sources:

```{r triangle-check}
ca_data <- reporting_data |>
  filter(location == "ca") |>
  cumulative_to_incremental(group_cols = c("reference_date", "location", "signal"))

reporting_triangle <- as_reporting_triangle(ca_data, delays_unit = "weeks")
print(reporting_triangle)
```

## Environment Requirements

This data source expects the machine running R to have:

- the `dataops_save` command on `PATH`, or a full command path supplied via `command`
- the relevant cfa-dataops catalog package installed
- Azure authentication and permissions for the backing storage

These requirements are only checked when this data source is used.
They are not required to install or use the rest of `nowcastNHSN`.
