---
title: "Using Internal cfa-dataops Data with nowcastNHSN"
output:
  rmarkdown::html_vignette:
    code_folding: show
vignette: >
  %\VignetteIndexEntry{Using Internal cfa-dataops Data with nowcastNHSN}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(nowcastNHSN)
library(baselinenowcast)
library(ggplot2)
library(dplyr)
```

## Introduction

The `nowcastNHSN` package supports several data sources for fetching vintaged NHSN hospital admissions data.
The [Getting Started](getting-started.html) vignette demonstrates the Delphi Epidata API, and the [Using Forecast Hub Data](hubdata-source.html) vignette demonstrates pulling target timeseries data from the forecast hub S3 buckets.

This vignette shows a third option, `cfa_dataops_source()`, which is intended for CDC-internal users.
It reads NHSN hospital admissions snapshots archived by the CDC's cfa-dataops tooling, giving access to the same week-by-week reporting history that the cfa-stf nowcasting and forecasting pipelines use internally.

Because this source relies on internal tooling and Azure storage, most users will prefer the Delphi or forecast hub sources.
For those who do have access, the interface returns data in the same reporting-triangle schema as the other sources, so the rest of the `nowcastNHSN` and `baselinenowcast` workflow is unchanged.

## Creating a cfa-dataops Source

Unlike the other sources, this one needs a small amount of setup outside R.
You will need the `cfa-dataops` Python package and the relevant public catalog installed in an environment that R can reach, and you will need Azure permissions for the underlying storage.

A minimal `uv`-based setup looks like this:

```{bash install-cfa-dataops, eval = FALSE}
uv venv /tmp/nowcastnhsn-cfa-dataops-venv --python 3.13
UV_CACHE_DIR=/tmp/uv-cache uv pip install \
  --python /tmp/nowcastnhsn-cfa-dataops-venv/bin/python \
  "cfa-dataops @ git+https://github.com/CDCgov/cfa-dataops.git@main" \
  "cfa-catalog-public @ git+https://github.com/CDCgov/cfa-catalog-pub.git"

export DATAOPS_SAVE=/tmp/nowcastnhsn-cfa-dataops-venv/bin/dataops_save
```

With that in place, `cfa_dataops_source()` creates a source object.
The `target` argument selects the disease signal, and `command` points to the `dataops_save` executable that R will call to fetch each snapshot.

```{r create-source, eval = FALSE}
source <- cfa_dataops_source(
  target = "covid",
  command = Sys.getenv("DATAOPS_SAVE", "dataops_save")
)
```

The available disease targets mirror the NHSN admissions columns used in the cfa-stf pipeline:

- `target = "covid"` for new confirmed COVID-19 admissions
- `target = "flu"` for new confirmed influenza admissions
- `target = "rsv"` for new confirmed RSV admissions

By default the source reads from the preliminary NHSN dataset (`public.stf.nhsn_hrd_prelim`).
Set `prelim = FALSE` to use the finalised dataset (`public.stf.nhsn_hrd`) instead.

## Fetching Reporting Data

The interface matches the other sources: pass Saturday week-ending dates for `reference_dates` and `report_dates`, and a vector of two-letter state abbreviations for `locations`.
Snapshots taken at different points within the same MMWR epiweek are mapped to the Saturday ending that week, so that `report_date` is always a Saturday and lines up with the forecast hub convention.

```{r fetch-data, eval = FALSE}
reporting_data <- fetch_reporting_data(
  source = source,
  reference_dates = seq(
    as.Date("2025-05-10"),
    as.Date("2025-08-30"),
    by = "week"
  ),
  report_dates = seq(
    as.Date("2025-08-16"),
    as.Date("2025-09-06"),
    by = "week"
  ),
  locations = c("ca", "ny")
)
```

```{r fetch-data-fixture, include = FALSE}
# Load pre-computed fixture to avoid requiring cfa-dataops and Azure access on
# every vignette build. To regenerate: run the chunk above in a cfa-dataops
# enabled environment and saveRDS() the result to
# vignettes/cfa-dataops-source/reporting_data.rds #nolint
reporting_data <- readRDS("cfa-dataops-source/reporting_data.rds")
```

For the rest of this vignette, the `reporting_data` shown below is loaded from a small pre-saved fixture rather than a live fetch.
This lets the vignette build without requiring cfa-dataops or Azure access, but the shape and meaning of the data are exactly what a live call would return.

```{r show-data}
head(reporting_data)
```

The returned data frame has the same columns as the Delphi and forecast hub sources:

- `reference_date`: the Saturday ending the week when admissions occurred
- `report_date`: the Saturday ending the MMWR epiweek when the snapshot was taken
- `location`: lowercase two-letter state abbreviation
- `count`: cumulative admissions for that reference date, as of that report date
- `signal`: the NHSN column name for the chosen target (e.g. `totalconfc19newadm` for COVID-19)

If more than one cfa-dataops snapshot falls in the same epiweek, the default behaviour is to keep the most recent snapshot for that `report_date` (`dedup = "latest"`).
Pass `dedup = "earliest"` to keep the first snapshot in each report week instead:

```{r dedup, eval = FALSE}
reporting_data <- fetch_reporting_data(
  source = source,
  reference_dates = "*",
  report_dates = seq(
    as.Date("2025-08-16"),
    as.Date("2025-09-06"),
    by = "week"
  ),
  locations = "ca",
  dedup = "earliest"
)
```

## Visualizing Reporting Delays

As with the other sources, counts for the same reference date change as later snapshots come in.
The plot below uses California COVID-19 admissions from the fixture to show how the cumulative counts at a handful of reference dates evolve across report dates.

```{r plot-delays, fig.width = 7, fig.height = 5, class.source = 'fold-hide'}
ca_data <- reporting_data |>
  filter(location == "ca")

all_report_dates <- sort(unique(ca_data$report_date))

selected_reports <- ca_data |>
  filter(report_date %in% all_report_dates)

ggplot(selected_reports, aes(x = reference_date, y = count, color = as.factor(report_date))) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  labs(
    title = "COVID-19 Hospital Admissions in California (cfa-dataops)",
    subtitle = "How reported counts change across snapshot vintages",
    x = "Reference Date (Week Ending)",
    y = "Confirmed Admissions",
    color = "Report Date"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom") +
  scale_color_brewer(palette = "Set1")
```

Each coloured line is a single snapshot (report date); each point on that line is the cumulative admissions reported for a given reference date as of that snapshot.
Recent reference dates tend to be incomplete on their first appearance and are typically revised upward in later snapshots, though downward revisions also happen.

## Preparing for Nowcasting

Because the returned data matches the common reporting-triangle schema, it plugs into the same pipeline as the other sources.
We convert the cumulative counts to incremental (new reports at each snapshot) and then build a reporting triangle:

```{r triangle-check}
ca_incremental <- reporting_data |>
  filter(location == "ca") |>
  cumulative_to_incremental(group_cols = c("reference_date", "location", "signal"))

reporting_triangle <- as_reporting_triangle(ca_incremental, delays_unit = "weeks")
print(reporting_triangle)
```

From here, the reporting triangle can be passed to `baselinenowcast::baselinenowcast()` in the same way as in the [Getting Started](getting-started.html) vignette.

## Environment Requirements

This source only checks its environment at the point of use; you can install and use the rest of `nowcastNHSN` without any of it.
When you do want to fetch live data through `cfa_dataops_source()`, the machine running R needs:

- the `dataops_save` command on `PATH`, or a full command path supplied via `command`
- the relevant cfa-dataops catalog package installed in the same Python environment
- Azure authentication and permissions for the backing storage
